### 2017-07-07 paper/conclu.tex
small edits
added Bradley Manna Sipma citations to a couple of points

%%\smallskip
\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied 
in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied 
in~\cite{HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17,MM16:proofrule-arxiv}. 
Supermartingales were also 
considered for other liveness and safety 
properties~\cite{CVS16:martingale-recurrence-persistence,BEFH16:doob}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work.

\noindent{\em Compositional a.s. termination proving.} 
A 
compositional rule for proving almost-sure termination was studied 
in~\cite{HolgerPOPL} under the uniform integrability assumption. 
In~\cite{MM05}, a soundness of the probabilistic variant rule is proved for 
programs with finitely many configurations.

@@

\noindent{\em Non-probabilistic programs.}
Termination analysis of non-probabilistic programs has also been extensively 
studied~\cite{PR04:transition-invariants, 
CPR06:terminator,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01, 
DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91,BMS05b,LJB01,KSTW10:compositional-transition-invariants,
 CPR11:termination-cacm}.
Ranking functions are at the heart of the termination analysis, and lexicographic 
ranking function has emerged as one of the most efficient and practical approaches
for termination analysis~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal}, 
being used e.g. in 
the prominent \texttt{T2} temporal prover~\cite{BCIKP16:T2}.
In this work we extend lexicographic ranking functions to probabilistic 
programs,
and present lexicographic RSMs for almost-sure termination analysis of probabilistic programs
with non-determinism. Theoretical complexity of synthesizing lexicographic 
ranking functions in non-probabilistic programs was studied 
in~\cite{BG13:integer-ranking,BG15:lexicographic-complexity}.

>>>>

%%\smallskip
\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied 
in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied 
in~\cite{HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17,MM16:proofrule-arxiv}. 
Supermartingales were also 
considered for other liveness and safety 
properties~\cite{CVS16:martingale-recurrence-persistence,BEFH16:doob}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work. 
%
%\noindent{\em Compositional a.s. termination proving.} 
A 
compositional rule for proving almost-sure termination was studied 
in~\cite{HolgerPOPL} under the uniform integrability assumption. 
In~\cite{MM05}, a soundness of the probabilistic variant rule is proved for 
programs with finitely many configurations.

@@

\noindent{\em Non-probabilistic programs.}
Termination analysis of non-probabilistic programs has also been extensively 
studied~\cite{PR04:transition-invariants, 
CPR06:terminator,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01, 
DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91,BMS05b,LJB01,KSTW10:compositional-transition-invariants,
 CPR11:termination-cacm}.
Ranking functions are at the heart of the termination analysis, and lexicographic 
ranking function has emerged as one of the most efficient and practical approaches
for termination 
analysis~\cite{DBLP:conf/cav/BradleyMS05,CSZ13,ADFG10:lexicographic,GMR15:rank-extremal},

being used e.g. in 
the prominent \texttt{T2} temporal prover~\cite{BCIKP16:T2}.
In this work we extend lexicographic ranking functions to probabilistic 
programs,
and present lexicographic RSMs for almost-sure termination analysis of probabilistic programs
with non-determinism. Theoretical complexity of synthesizing lexicographic 
ranking functions in non-probabilistic programs was studied 
in~\cite{BG13:integer-ranking,BG15:lexicographic-complexity}.

### 2017-07-07 paper/introduction.tex
Several small changes
added Bradley Manna Sipma citations to a couple of points

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While an impressive set of theoretical results related to RSMs has been
established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17},
 for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. This is because to prove termination, a witness in the form of a 
ranking function has to be computed: to do this automatically, ranking 
functions 
of a restricted form (such as linear ranking functions) have to be considered, 
and 1-dimensional ranking functions of a restricted type can only prove 
termination of a limited class of programs. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely 
studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2}. 
Algorithmic approaches for linear lexicographic ranking functions allow the 
termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

\smallskip\noindent{\em Our contributions.} 
In this work our main contributions range from theoretical foundations of lexicographic
RSMs, to algorithmic approaches for them, to experimental results
showing their applicability to programs.
We describe our main contributions below:
\begin{enumerate}

@@

\item {\em Asymptotic bounds.} 
In general, the existence of lexicographic RSMs does not imply positive
termination. 
In other words, we present an example (Example~\ref{ex:infinite-time}) where a 
lexicographic RSM exists ensuring
almost-sure termination, yet the expected termination time is infinite.
We then present a natural restriction under which the lexicographic 
RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 
time can be derived from them (Theorem~\ref{thm:runtime-bound} and 
Corollary~\ref{col:runtime-progs}).

@@

\end{enumerate}

>>>>

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While an impressive set of theoretical results related to RSMs has been
established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17},
 for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. This is because to prove termination, a witness in the form of a 
ranking function has to be computed: to do this automatically, ranking 
functions 
of a restricted form (such as linear ranking functions) have to be considered, 
and 1-dimensional ranking functions of a restricted type can only prove 
termination of a limited class of programs. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely 
studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2, 
DBLP:conf/cav/BradleyMS05}.

Algorithmic approaches for linear lexicographic ranking functions allow the 
termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

\smallskip\noindent{\em Our contributions.} 
In this work our main contributions range from theoretical foundations of lexicographic
RSMs, to algorithmic approaches for them, to experimental results
showing their applicability to programs.
We describe our main contributions below:
\begin{compactenum}

@@

\item {\em Asymptotic bounds.} 
In general, the existence of lexicographic RSMs does not imply positive
termination. 
In other words, we present an example (Example~\ref{ex:infinite-time}) where a 
lexicographic RSM exists ensuring
almost-sure termination, yet the expected termination time is infinite.
We then present a natural restriction under which the lexicographic 
RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 
time can be derived from them (Theorem~\ref{thm:runtime-bound} and 
Corollary~\ref{col:runtime-progs}).
Moreover, the asymptotic bounds we obtain establish polynomial bounds on the expected termination 
time (see Remark~\ref{rem:poly}).

@@

\end{compactenum}
